---
title: "Lab 12 - Smoking during pregnancy"
author: "Hannah Crawley"
date: "4/10/2025"
output: github_document
---

### Load packages and data

```{r load-packages, message=FALSE}
library(tidyverse) 
library(tidymodels)
library(dplyr)
library(openintro)
```

```{r}
data("ncbirths")
```

```{r}
boxplot(ncbirths$fage)
boxplot(ncbirths$mage)
boxplot(ncbirths$weeks)
boxplot(ncbirths$visits)
boxplot(ncbirths$weight)
```

```{r}

Q1 <- quantile(ncbirths$weight, 0.25)
Q3 <- quantile(ncbirths$weight, 0.75)
IQR <- Q3 - Q1

ncbirths_outliers <- ncbirths %>%
  filter(weight < (Q1 - 1.5 * IQR) | weight > (Q3 + 1.5 * IQR))

```

```{r}
summary(ncbirths_outliers)
```
## Part 1: Baby Weights 

### Exercise 1
> What are the cases in this data set? How many cases are there in our sample?

There are 1000 observations (each observation is one mother/pregnancy) and 13 variables 

### Exercise 2
> Create a filtered data frame called ncbirths_white that contains data only from White mothers. Then, calculate the mean of the weights of their babies.

```{r}
ncbirths_white <- ncbirths %>%
  filter(whitemom == "white") 
```

```{r}
ncbirths_notwhite <- ncbirths %>%
  filter(whitemom == "not white") 
```

```{r}
mean(ncbirths_white$weight)
```

```{r}
mean(ncbirths_notwhite$weight)
```

### Exercise 3
> Are the criteria necessary for conducting simulation-based inference satisfied? Explain your reasoning.


### Exercise 4
> Run the appropriate hypothesis test, visualize the null distribution, calculate the p-value, and interpret the results in the context of the data and the hypothesis test.

Null - baby weight = 7.43
Alternative - weight does not = 7.43

```{r}
boot_dist <- ncbirths_white %>%
  specify(response = weight) %>%
  generate(reps = 1000, type = "bootstrap") %>%
  calculate(stat = "mean")
```

```{r}
ggplot(boot_dist, aes(x = stat)) +
  geom_histogram(binwidth = 0.01) +
  labs(title = "Null distribution")
```
```{r}
get_p_value(boot_dist, obs_stat = 7.43, direction = "two-sided")
```


```{r}
t.test(ncbirths_white$weight, mu = 7.43)
```

The mean of the current sample is 7.25 as compared to the 7.43 reported in 1995 -- the t-test is significant (p = 0.0009) suggesting there is a sig difference between the means. We can infer that the average weight of babies has changed since 1995.

## Part 2: Baby weight vs smoking 

### Exercise 5
> Make side-by-side box plots displaying the relationship between habit and weight. What does the plot highlight about the relationship between these two variables?

```{r}
ggplot(ncbirths_white, aes(x = weight, y = habit)) +
  geom_boxplot(outlier.shape = NA, fill = "lightgray") +
  geom_jitter(width = 0.2, size = 2, alpha = 0.6, color = "black") +
  labs(x = "Weight", y = "Habit") 
```
From these box plots - non smokers' babies tend to have higher birth weights on average than smokers' babies suggesting there is a potential neg effect of smoking on birth weight 

### Exercise 6
> Before continuing, create a cleaned version of the dataset by removing any rows with missing values for habit or weight. Name this version ncbirths_clean.

### Exercise 7
> Calculate the observed difference in means between the baby weights of smoking and non-smoking mothers.

### Exercise 8
> Write the hypotheses for testing if the average weights of babies born to smoking and non-smoking mothers are different.

H0: _________ (μ1 = μ2)

HA: _________ (μ1 ≠ μ2)

### Exercise 9
> Run the appropriate hypothesis test, calculate the p-value, and interpret the results in context of the data and the hypothesis test.

### Exercise 10
> Construct a 95% confidence interval for the difference between the average weights of babies born to smoking and non-smoking mothers.

## Part 3: Mother's age vs baby weight 

### Exercise 11
> First, a non-inference task: Determine the age cutoff for younger and mature mothers. Use a method of your choice, and explain how your method works.

### Exercise 12
> Conduct a hypothesis test evaluating whether the proportion of low birth weight babies is higher for mature mothers. Use  
α = 0.05. State the hypotheses, Verify the conditions, Run the test and calculate the p-value, State your conclusion within context of the research question

### Exercise 13
> Calculate a confidence interval for the difference between the proportions of low birth weight babies between mature and younger mothers. Interpret the interval in the context of the data and explain what it means.

